Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups

### 2020-04-02 llncs2e/paper77.tex
corrected assumptions

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups\thanks{The work is partially supported by Siemens Power and Gas. The code is published on GitHub: \href{https://github.com/karhunenloeve/PHIA}{https://github.com/karhunenloeve/PHIA}.}}
%
\titlerunning{Invertible Resnets up to Homology Equivalence}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an elusive number of scientific disciplines. Nevertheless, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After processing through a layer, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer for an invertible residual net. We consider the data domain as abelian connected matrix group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

\subsection{Simplicial Complexes and Homology Theory}
\label{simcomplexhomtheory}
Simplicial homology is studied by $\Delta$-complexes whose simplices are uniquely determined by their vertices. Thus, every $n$-simplex has $(n+1)$ distinct vertices and no other $n$-simplex has the same set of vertices. A \emph{simplicial complex} $K$ is a collection of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. This is a unique combinatorial description of vertices together with a collection of sets $\{\Delta^k_1, \cdots, \Delta^k_m\}$ of $k$-simplices, which are $(k+1)$ element subsets of $K$ \citemain[p.~107]{hatcher2002algebraic}. Further, we require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $\Delta^k \subset K$. Simplicial complexes have a natural order of simplices, which is illustrated in Fig. \ref{complexes}, which plays a central role in their geometric realization \citemain[p.~14]{boissonnat2018geometric}. If we choose a total order for all corners which are located in any simplex of $K$, we get an order for each single $k$-simplex.

A \textit{singular $k$-simplex} is defined as a continuous map $\sigma: \Delta^k \rightarrow X$ into a topological space $X$. Intuitively, every simplicial complex $K$ has some connected components. These form a vector space $C_k(X)$, which is called \textit{$k$-th chain group}, with coefficients in $\mathbb{Z}$ and elements called $k$-\textit{chains}. The $k$-chains are finite sums $\sum_{i=0}^{k} z_i \sigma_i$ with $z_i \in \mathbb{Z}$ \citemain[p.~108]{hatcher2002algebraic}. Having connected components defined as a group implies also the definition of a boundary. As the boundaries of a $k$-simplex are also $(k-1)$-simplices, we can remove all other simplices to project into the lower chain group by the $k$-th boundary operator $\partial_k: C_k(X) \rightarrow C_{k-1}(X)$:
\begin{equation}
    \partial_k(\sigma) = \sum_{i=0}^{k} (-1)^{i} \sigma \bigg\vert {[v_0,\cdots,\hat{v}_i, \cdots, v_k]}.
\end{equation}
Observe that the composition of boundary maps is zero, as a boundary has no boundary by itself, i.e. $\partial_i\partial_{i+1} = 0$ \citemain[pp.~108--109]{hatcher2002algebraic}. Thus, $\ker \; \partial_i$ corresponds to the generating elements of the $i$-th boundary group and $\Ima \; \partial_{i+1}$ for the $i$-th chain group. The singular homology group is the quotient space
\begin{equation}
    H_k(X) = \frac{\ker \left(\partial_k: C_k\left(X\right) \rightarrow C_{k-1}\left(X\right)\right)}{\Ima \left(\partial_{k+1}:C_{k+1}\left(X\right) \rightarrow C_{k}\left(X\right)\right)}.
\end{equation}
The $k$-th singular homology group is isomorphic to the $n$-th simplicial homology group for a trianguliable metric space, i.e. $H_k(X) \simeq H^{\Delta}_k(X)$ for all $k \in \mathbb{Z}$, see Appx. \ref{simplicialsingularequivalence}, thus we can transfer results from one homology theory to the other.

@@

\subsection{Persistence Diagrams}
Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}. The Betti numbers are defined by $\rank(H_k(X))$. We write $H^{\omega_i}_k = H_k(K^{\omega_i}(X))$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\omega_i$. Then $H^{\omega_{i}}_{k} \rightarrow H^{\omega_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, see Appx. \ref{persistenthomologyproofs} \citemain{lume,edelsbrunner2008persistent}:
\begin{align}
0 = H^{\omega_{0}}_k \rightarrow H^{\omega_{1}}_k \rightarrow \cdots \rightarrow H^{\omega_{n}}_k \rightarrow H^{\omega_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\omega_i}(X)$ or appear before and die after spanning $K^{\omega_{i+1}}(X)$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\omega_{i},\omega_{j}}_k := (\beta_k^{\omega_{i},\omega_{j-1}} - \beta_k^{\omega_i,\omega_{j}})-(\beta_k^{\omega_{i-1},\omega_{j-1}}-\beta_k^{\omega_{i-1},\omega_{j}})
\end{equation}
for the $k$-th homology group and index pairs $(\omega_i,\omega_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$ \citemain{edelsbrunner2008persistent}. The Euclidean space is extended, as the first connected component of our filtration is not merged with any other critical value. We assign the value $\infty$ to it, corresponding to the second coordinate $w_{j+1}$ in the persistence diagram. The first term counts elements born in $K^{\omega_{j-1}}(X)$ and vanished entering $K^{\omega_{j}}(X)$, while the second term counts the representatives of homology classes before $K^{\omega_{j}}(X)$ and vanishing at $K^{\omega_{j}}(X)$. The index pairs denote the \emph{birth} and \emph{death} of a homology class. The $k$-th persistence diagram is thus
\begin{equation}
\mathfrak{P}_k(K(X)) := \left\{(\omega_i, \omega_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\omega_{i},\omega_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

@@

\subsection{Counting Betti Numbers}
In the following, the homology groups are computed over $\mathbb{Z}$. Based on our assumption to operate on a connected commutative Lie group, there is an isomorphism for the homology groups of the data space $G$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$, see \ref{decomposition}. The homology groups of this product space are equal to the homology groups of the sought manifold. The neural net should be able to preserve all structures within this space. Since the input domain is a set of points to be described by one or more compact manifolds, all homology classes should be preserved in our learning task. Künneth's theorem states that the $k$-th singular homology group of a topological product space $X \times Y$ is isomorphic to the product of the respective $k$-th homology groups of $X$ and $Y$, see \citemain[p.~268]{hatcher2002algebraic} and Appx. \ref{a:groups}:
\begin{align}
    H_k(X \times Y) \simeq \bigoplus_{i+j=k}H_i(X) \otimes H_j(Y).
\end{align}
This applies to all fields, since modules over a field are always free. If we apply the theorem to the decomposition, we get for the space we are looking for
\begin{align}
	\label{homologykunneth}
    &H_k(\mathcal{S}^{1}_{1}\times\cdots\times \mathcal{S}^{1}_{q}\times\mathbb{R}^p) \simeq \\
    \simeq \bigoplus_{i_{1} + \cdots + i_{r}=k} &H_{i_{1}}(\mathcal{S}^{1}_{1}) \otimes \cdots \otimes H_{i_{\hat{r}}}(\mathcal{S}^{1}_{q}) \otimes H_{i_{r}}(\mathbb{R}^{p}).
\end{align}
We need to know how to count in order to derive the dimension of the sought manifold from the representatives of homology classes within the mean persistence landscapes. Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, the homology group $H_k(\mathbb{T}^{q}) \simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}}$ and $H_0(\mathbb{R}^p)\simeq\mathbb{Z}^{p}$, where $p$ indicates the number of connected components. We denote the amount of representatives from homology classes of the $k$-th homology group appearing in the mean persistence landscape $\bar{\mathfrak{L}}$ by $\sharp^{H}_{k}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X)))$. Thus, we get the following relation for the representatives of the zeroth homology group:
\begin{align}
	\label{countingsimple}
	\sharp_{0}^{H}(\bar{\mathfrak{L}}(K(X))) \approx p.
\end{align}
This is the number of connected components to be represented by the innermost layer of the neural net, corresponding exactly to the number of elements from the zeroth homology group. We only need to consider the homology groups which are represented in the mean persistence landscape. Assuming that our estimate is good, higher homology groups are trivial. The required number of factors of $1$-spheres can be determined for the $k$-th homology group by
\begin{align}
	\label{countingequation}
	\sharp_{k}^{H}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X))) \approx \prod_{i=1}^{k} \frac{q-k+i}{i}
\end{align}

Note, that we assume to work on a smooth manifold, whose de Rham cohomology induces an isomorphism into the simplicial homology with coefficients in $\mathbb{R}$, see Appx. \ref{simplicialsingularequivalence}. We can still work with coefficients in $\mathbb{Z}$ because, according to the Universal Coefficient Theorem, the integral homology groups determine the homology groups over every field \cite{gruenberg1968universal}. Thus, these results can be applied without restriction to any homology theory, see Appx. \ref{simplicialsingularequivalence}.

@@

\subsection{Deep Learning}
\input{figure2.tex}
Neural nets are used in practice with a different number of perceptrons per layer. However, a layer, from the perspective of a matrix group, can also represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map remains constant. Thus, the dimension of the data manifold is the dimension of the smallest possible bottleneck layer of the neural net  $\dim M = \min_l \dim \textbf{x}^{(l)}$ \citemain{HauserR17}. Our contribution provides the theory of how this dimension can be determined for all neural nets that are a composition of invertible layers, i.e. $\textbf{x}^{(l)} \in \text{GL}(n,\mathbb{R}) := \{\varphi: \mathbb{R}^n \rightarrow \mathbb{R}^n \; \vert \; \det \varphi \neq 0 \}$. The concatenation of such maps is called a deep neural net:
\begin{equation}
    \textbf{x}^{(L)}(M) = \left(\varphi^{(l+1)} \circ \varphi^{(l)} \cdots \circ \varphi^{(1)} \circ \varphi^{(0)} \circ \textbf{x}^{(0)}\right)(M).
\end{equation}
Activation functions restrict to a subgroup of Euclidean space. Continuous homomorphisms between Lie groups are automatically smooth. Thus, a second countable locally Euclidean group has exactly one differentiable structure, turning it into a Lie group. We need to show that $\textbf{x}^{(l+1)}$ has the topology of $G$ and is therefore an embedded closed Lie subgroup $\textbf{x}^{(l)} := H \subseteq G$. Cartans Theorem of Closed Subgroups states that a subgroup $H$ of a Lie group $G$ is an embedded Lie subgroup, if it is a closed subgroup \citemain{cartan1952theorie}. We construct our neural nets such that they are a subset of $\text{GL}(n,\mathbb{R})$ and according to the proof in Appx. \ref{liegroups}, this is a closed subgroup of $\text{GL}(n,\mathbb{C})$, which is itself a Lie group. It follows that a layer in the neural net is also a Lie group. Due to invertibility, the subsequent smaller layers are Lie subgroups, as $\sigma^{(l)}$ is smooth.

@@

\subsection{Invertible Architecture}
\label{subsec:invertibleneuralnets}
The group $\text{GL}(n,\mathbb{R})$ is the group of invertible matrices, thus we have to equip our neural net with an invertible architecture. The input vector is given by $\varphi^{(l)} := \textbf{x}^{(l)} = (\textbf{u}^{(l)}_{1},\textbf{u}^{(l)}_{1})$, splitting the $l$-th layer into $\textbf{u}^{(l)}_1$ and $\textbf{u}^{(l)}_2$ \citemain{DinhSB17,ArdizzoneKRK19}:
\begin{align}
	\textbf{u}^{(l)}_{1} &= \left(\textbf{v}^{(l)}_1 - \psi_2\left(\textbf{u}^{(l)}_2\right)\right) \cdot \exp\left(-\xi_2\left(\textbf{u}^{(l)}_2\right)\right), \\
	\textbf{u}^{(l)}_{2} &= \left(\textbf{v}^{(l)}_2 - \psi_2\left(\textbf{v}^{(l)}_1\right)\right) \cdot \exp\left(-\xi_1\left(\textbf{v}^{(l)}_1\right)\right).
\end{align}
The two input vectors $\textbf{v}^{(l)}_1$ and $\textbf{v}^{(l)}_2$ of the layer ensure that the output is trivially invertible. The multiplication is to be understood elementwise, as in Eq. \ref{neuralnet}:
\begin{align}
	\textbf{v}_1 = \textbf{u}^{(l)}_{1} \cdot \exp\left(\xi_2\left(\textbf{u}^{(l)}_{2}\right) + \psi_2\left(\textbf{u}^{(l)}_{2}\right)\right), \\
	\textbf{v}_1 = \textbf{u}^{(l)}_{1} \cdot \exp\left(\xi_2\left(\textbf{u}^{(l)}_{2}\right) + \psi_2\left(\textbf{u}^{(l)}_{2}\right)\right).
\end{align} 
We need a completely invertible neural net, therefore all layers are invertible -- except the very first one, see Fig. \ref{architecture} and Sect. \ref{sec:numericalexperiments}. The functions $\psi_i,\xi_j: \textbf{u}^{(l)}\left(M\right) \rightarrow \left(\xi_j \circ \textbf{u}^{(l)}\right)\left(M\right)$, $\textbf{u}^{(l)} \mapsto \sigma\left(W^{(l+1)} \cdot \textbf{u}^{(l)} + \textbf{b}^{(l+1)}\right)$ are implemented as neural subnets by themselves. These nets do not have to be invertible in general.

\section{Numerical Experiments}
\label{sec:numericalexperiments}
\input{figure3.tex}
We train an autoencoder and use as input heavily noisy images and map them to their noiseless original. It is not relevant whether the mean persistence landscapes are computed over the noisy data or over the originals, since the primer can deal with noise. In principle, the persistence landscapes should be calculated for the data manifold. We use invertible $\mathcal{C}^1$-nets and isolate the innermost layer of the autoencoder, see Sect. \ref{resnet} and Fig. \ref{architecture}. To test if our prediction for the embedding dimension, we parameterize the innermost layer using two perceptrons. The nets are trained and evaluated according to the caption of Fig. \ref{costfunctions}.

>>>>

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Invertible Resnets up to Homology Equivalence for Abelian Lie Groups}
%
\titlerunning{Invertible Resnets up to Homology Equivalence}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an elusive number of scientific disciplines. Nevertheless, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After processing through a layer, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer for an invertible residual net. We consider the data domain as abelian connected matrix group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

\subsection{Simplicial Complexes and Homology Theory}
\label{simcomplexhomtheory}
Simplicial homology is studied by $\Delta$-complexes whose simplices are uniquely determined by their vertices. Thus, every $n$-simplex has $(n+1)$ distinct vertices and no other $n$-simplex has the same set of vertices. A \emph{simplicial complex} $K$ is a collection of simplices, such that any face of a simplex of $K$ is a simplex of $K$ and the intersection of any two simplices of $K$ is either empty or a common face of both \cite[p.~11]{boissonnat2018geometric}. This is a unique combinatorial description of vertices together with a collection of sets $\{\Delta^k_1, \cdots, \Delta^k_m\}$ of $k$-simplices, which are $(k+1)$ element subsets of $K$ \citemain[p.~107]{hatcher2002algebraic}. Further, we require every $(k+1)$-element, with $k < n$ to be a $k$-simplex of $\Delta^k \subset K$. Simplicial complexes have a natural order of simplices, which is illustrated in Fig. \ref{complexes}. It plays a central role in their geometric realization \citemain[p.~14]{boissonnat2018geometric}. If we choose a total order for all corners which are located in any simplex of $K$, we get an order for each single $k$-simplex.

A \textit{singular $k$-simplex} is defined as a continuous map $\sigma: \Delta^k \rightarrow X$ into a topological space $X$. Intuitively, every simplicial complex $K$ has some connected components. These form a vector space $C_k(X)$, which is called \textit{$k$-th chain group}, with coefficients in $\mathbb{Z}$ and elements called $k$-\textit{chains}. The $k$-chains are finite sums $\sum_{i=0}^{k} z_i \sigma_i$ with $z_i \in \mathbb{Z}$ \citemain[p.~108]{hatcher2002algebraic}. Having connected components defined as a group implies also the definition of a boundary. As the boundaries of a $k$-simplex are also $(k-1)$-simplices, we can remove all other simplices to project into the lower chain group by the $k$-th boundary operator $\partial_k: C_k(X) \rightarrow C_{k-1}(X)$:
\begin{equation}
    \partial_k(\sigma) = \sum_{i=0}^{k} (-1)^{i} \sigma \bigg\vert {[v_0,\cdots,\hat{v}_i, \cdots, v_k]}.
\end{equation}
Observe that the composition of boundary maps is zero, as a boundary has no boundary by itself, i.e. $\partial_i\partial_{i+1} = 0$ \citemain[pp.~108--109]{hatcher2002algebraic}. Thus, $\ker \; \partial_i$ corresponds to the generating elements of the $i$-th boundary group and $\Ima \; \partial_{i+1}$ for the $i$-th chain group. The singular homology group is the quotient space
\begin{equation}
    H_k(X) = \frac{\ker \left(\partial_k: C_k\left(X\right) \rightarrow C_{k-1}\left(X\right)\right)}{\Ima \left(\partial_{k+1}:C_{k+1}\left(X\right) \rightarrow C_{k}\left(X\right)\right)}.
\end{equation}
The $k$-th singular homology group is isomorphic to the $k$-th simplicial homology group for a trianguliable metric space, i.e. $H_k(X) \simeq H^{\Delta}_k(X)$ for all $k \in \mathbb{Z}$, see Appx. \ref{simplicialsingularequivalence}, thus we can transfer results from one homology theory to the other.

@@

\subsection{Persistence Diagrams}
Recording the \emph{Betti numbers} of the $k$-th homology group along the filtration, we obtain the \emph{$k$-dimensional persistence diagram}. The Betti numbers $\beta_k$ are defined by $\rank(H_k(X))$. We write $H^{\omega_i}_k = H_k(K^{\omega_i}(X))$ as $k$-th homology group on the simplicial complex $K$ with parametrization $\omega_i$. Then $H^{\omega_{i}}_{k} \rightarrow H^{\omega_{i+1}}_k$ induces a sequence of homomorphisms on the filtration, see Appx. \ref{persistenthomologyproofs} \citemain{lume,edelsbrunner2008persistent}:
\begin{align}
0 = H^{\omega_{0}}_k \rightarrow H^{\omega_{1}}_k \rightarrow \cdots \rightarrow H^{\omega_{n}}_k \rightarrow H^{\omega_{n+1}}_k = 0.
\end{align}
The image of each homomorphism consists of all $k$-dimensional homology classes which are born in $K^{\omega_i}(X)$ or appear before and die after spanning $K^{\omega_{i+1}}(X)$. Tracking the Betti numbers on the filtration results into a multiplicity
\begin{equation}
\mu^{\omega_{i},\omega_{j}}_k := (\beta_k^{\omega_{i},\omega_{j-1}} - \beta_k^{\omega_i,\omega_{j}})-(\beta_k^{\omega_{i-1},\omega_{j-1}}-\beta_k^{\omega_{i-1},\omega_{j}})
\end{equation}
for the $k$-th homology group and index pairs $(\omega_i,\omega_{j+1}) \in \overline{\mathbb{R}^2} := \mathbb{R}^2 \cup \infty$ with indices $i \leq j$ \citemain{edelsbrunner2008persistent}. The Euclidean space is extended, as the first connected component of our filtration is not merged with any other critical value. We assign the value $\infty$ to it, corresponding to the second coordinate $w_{j+1}$ in the persistence diagram. The first term counts elements born in $K^{\omega_{j-1}}(X)$ and vanished entering $K^{\omega_{j}}(X)$, while the second term counts the representatives of homology classes before $K^{\omega_{j}}(X)$ and vanishing at $K^{\omega_{j}}(X)$. The index pairs denote the \emph{birth} and \emph{death} of a homology class. The $k$-th persistence diagram is thus
\begin{equation}
\mathfrak{P}_k(K(X)) := \left\{(\omega_i, \omega_{j+1}) \in \overline{\mathbb{R}^2} \; \bigg\vert \; \mu^{\omega_{i},\omega_{j+1}}_k = 1 \; \text{for all} \; i,j \in I \right\}.
\end{equation}

@@

\subsection{Counting Betti Numbers}
In the following, the homology groups are computed over $\mathbb{Z}$. Based on our assumption to operate on a connected commutative Lie group, there is an isomorphism for the homology groups of the data space $G$, so that $H_k(G) \simeq H_k(\mathbb{R}^p \times \mathcal{S}^{1}_{1}\times\cdots\times\mathcal{S}^{1}_{q})$ with $p+q = \dim G$, see \ref{decomposition}. The homology groups of this product space are equal to the homology groups of the sought manifold. The neural net should be able to preserve all structures within this space. Since the input domain is a set of points to be described by one or more compact manifolds, all homology classes should be preserved in our learning task. Künneth's theorem states that the $k$-th singular homology group of a topological product space $X \times Y$ is isomorphic to the product of the respective $k$-th homology groups of $X$ and $Y$, see \citemain[p.~268]{hatcher2002algebraic} and Appx. \ref{a:groups}:
\begin{align}
    H_k(X \times Y) \simeq \bigoplus_{i+j=k}H_i(X) \otimes H_j(Y).
\end{align}
This applies to all fields, since modules over a field are always free. If we apply the theorem to the decomposition, we get for the space we are looking for
\begin{align}
	\label{homologykunneth}
    &H_k(\mathcal{S}^{1}_{1}\times\cdots\times \mathcal{S}^{1}_{q}\times\mathbb{R}^p) \simeq \\
    \simeq \bigoplus_{i_{1} + \cdots + i_{r}=k} &H_{i_{1}}(\mathcal{S}^{1}_{1}) \otimes \cdots \otimes H_{i_{\hat{r}}}(\mathcal{S}^{1}_{q}) \otimes H_{i_{r}}(\mathbb{R}^{p}).
\end{align}
We need to know how to count in order to derive the dimension of the sought manifold from the representatives of homology classes within the mean persistence landscapes. Note, in Eq. \ref{homologykunneth} terms remain only for indices $i_{j} \in \{0,1\}$. Thus, the homology group $H_k(\mathbb{T}^{q}) \simeq H_k(\mathcal{S}^{1}_{1} \times \cdots \times \mathcal{S}^{1}_{q}) \simeq \mathbb{Z}^{{q\choose{k}}}$ and $H_0(\mathbb{R}^p)\simeq\mathbb{Z}^{p}$, where $p$ indicates the number of connected components. We denote the amount of representatives from homology classes of the $k$-th homology group appearing in the mean persistence landscape $\bar{\mathfrak{L}}$ by $\sharp^{H}_{k}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X)))$. Thus, we get the following relation for the representatives of the zeroth homology group:
\begin{align}
	\label{countingsimple}
	\sharp_{0}^{H}(\bar{\mathfrak{L}}(K(X))) \approx p.
\end{align}
This is the number of connected components to be represented by the innermost layer of the neural net, corresponding exactly to the number of elements from the zeroth homology group. We only need to consider the homology groups which are represented in the mean persistence landscape. Assuming that our estimate is good, higher homology groups are trivial. The required number of factors of $1$-spheres can be determined for the $k$-th homology group by
\begin{align}
	\label{countingequation}
	\sharp_{k}^{H}(\bar{\mathfrak{L}}(\mathfrak{P}(K(X))) \approx \prod_{i=1}^{k} \frac{q-k+i}{i}.
\end{align}
Note, that we assume to work on a smooth manifold, whose de Rham cohomology induces an isomorphism into the simplicial homology with coefficients in $\mathbb{R}$, see Appx. \ref{simplicialsingularequivalence}. We can still work with coefficients in $\mathbb{Z}$ because, according to the Universal Coefficient Theorem, the integral homology groups determine the homology groups over every field \cite{gruenberg1968universal}. Thus, these results can be applied without restriction to any homology theory, see Appx. \ref{simplicialsingularequivalence}.

@@

\subsection{Deep Learning}
\input{figure2.tex}
Neural nets are used in practice with a different number of perceptrons per layer. However, a layer, from the perspective of a matrix group, can also represent the manifold of the previous one. Manifolds can always be immersed and submersed as long as the rank of the Jacobian of the respective map remains constant. Thus, the dimension of the data manifold is the dimension of the smallest possible bottleneck layer of the neural net  $\dim M = \min_l \dim \textbf{x}^{(l)}$ \citemain{HauserR17}. Our contribution provides the theory of how this dimension can be determined for all neural nets that are a composition of invertible layers acting on an abelian Lie group as domain of the data. The concatenation of such maps is called a deep neural net:
\begin{equation}
    \textbf{x}^{(L)}(M) = \left(\varphi^{(l+1)} \circ \varphi^{(l)} \cdots \circ \varphi^{(1)} \circ \varphi^{(0)} \circ \textbf{x}^{(0)}\right)(M).
\end{equation}
Activation functions restrict to a subgroup of Euclidean space. Continuous homomorphisms between Lie groups are automatically smooth. Thus, a second countable locally Euclidean group has exactly one differentiable structure, turning it into a Lie group. We need to show that $\textbf{x}^{(l+1)}$ has the topology of $G$ and is therefore an embedded closed Lie subgroup $\textbf{x}^{(l)} := H \subseteq G$. Cartans Theorem of Closed Subgroups states that a subgroup $H$ of a Lie group $G$ is an embedded Lie subgroup, if it is a closed subgroup \citemain{cartan1952theorie}. We construct our neural nets such that they are a subset of $\text{GL}(n,\mathbb{R})$ and according to the proof in Appx. \ref{liegroups}, this is a closed subgroup of $\text{GL}(n,\mathbb{C})$, which is itself a Lie group. It follows that a layer in the neural net is also a Lie group. Due to invertibility, the subsequent smaller layers are Lie subgroups, as $\sigma^{(l)}$ is smooth.

@@

\subsection{Invertible Architecture}
\label{subsec:invertibleneuralnets}
The group $\text{GL}(n,\mathbb{R})$ is the group of invertible matrices, thus we have to equip our neural net with an invertible architecture. The input vector is given by $\varphi^{(l)} := \textbf{x}^{(l)} = (\textbf{u}^{(l)}_{1},\textbf{u}^{(l)}_{1})$, splitting the $l$-th layer into $\textbf{u}^{(l)}_1$ and $\textbf{u}^{(l)}_2$ \citemain{DinhSB17,ArdizzoneKRK19}:
\begin{align}
	\textbf{u}^{(l)}_{1} &= \left(\textbf{v}^{(l)}_1 - \psi_2\left(\textbf{u}^{(l)}_2\right)\right) \cdot \exp\left(-\xi_2\left(\textbf{u}^{(l)}_2\right)\right), \\
	\textbf{u}^{(l)}_{2} &= \left(\textbf{v}^{(l)}_2 - \psi_2\left(\textbf{v}^{(l)}_1\right)\right) \cdot \exp\left(-\xi_1\left(\textbf{v}^{(l)}_1\right)\right).
\end{align}
The two input vectors $\textbf{v}^{(l)}_1$ and $\textbf{v}^{(l)}_2$ of the layer ensure that the output is trivially invertible. The multiplication is to be understood elementwise, as in Eq. \ref{neuralnet}:
\begin{align}
	\textbf{v}_1 = \textbf{u}^{(l)}_{1} \cdot \exp\left(\xi_2\left(\textbf{u}^{(l)}_{2}\right) + \psi_2\left(\textbf{u}^{(l)}_{2}\right)\right), \\
	\textbf{v}_1 = \textbf{u}^{(l)}_{1} \cdot \exp\left(\xi_2\left(\textbf{u}^{(l)}_{2}\right) + \psi_2\left(\textbf{u}^{(l)}_{2}\right)\right).
\end{align} 
We operate on the Lie group $\text{GL}(n,\mathbb{R})$, therefore all layers are invertible -- except the very first one, see Fig. \ref{architecture} and Sect. \ref{sec:numericalexperiments}. The functions $\psi_i,\xi_j: \textbf{u}^{(l)}\left(M\right) \rightarrow \left(\xi_j \circ \textbf{u}^{(l)}\right)\left(M\right)$, $\textbf{u}^{(l)} \mapsto \sigma\left(W^{(l+1)} \cdot \textbf{u}^{(l)} + \textbf{b}^{(l+1)}\right)$ are implemented as neural subnets by themselves. These nets do not have to be invertible in general.

\section{Numerical Experiments}
\label{sec:numericalexperiments}
\input{figure3.tex}
We train an autoencoder and use as input heavily noisy images and map them to their noiseless original. It is not relevant whether the mean persistence landscapes are computed over the noisy data or over the originals, since the primer can deal with noise. In principle, the persistence landscapes should be calculated for the data manifold. We use invertible $\mathcal{C}^1$-nets and isolate the innermost layer of the autoencoder, see Sect. \ref{resnet} and Fig. \ref{architecture}. To test our prediction for the embedding dimension, we parameterize the innermost layer using a multiple of $2$ perceptrons. The nets are trained and evaluated according to Fig. \ref{costfunctions}.
